MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments

### 2022-05-19 main/MSstatsPTM_Main.tex
changes to spike in sections

\todo{This section would benefit from more biology/technology background, and more extensive references}
Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of bimolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen}. Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale \todo{would be good to have a reference}.

@@

The technological challenges in PTM identification and quantification challenge the downstream statistical analyses. Yet frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. In particular, methods such as $t$-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~\cite{Ritchie_15a}. Such methods do not fully account for all the sources of uncertainty and variation in the experiment. As the result, they are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

The datasets in this manuscript  benchmarked the proposed approach in situations with known ground truth, and represented a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real-world experimental settings with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, PTM types, experimental designs and acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments.

@@

The simulation represented an idealistic case. 24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation. In each dataset, 1000 proteins had 10 unmodified features per protein. Each of the 1000 proteins had one PTM, each PTM represented by 10 modified features. The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no difference between conditions. Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein. Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All differential PTMs were generated with an expected log fold change of 0.75 between conditions.

@@

\todo{Devon: I moved the figure to main and tried to adjust this section to be clearer. I still think this section could benefit from Erik and Tsung-Heng's insight. Will ask them to look in email.}

@@

\medskip \noindent {\bf Experimental Design:} 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark. An overview of the experimental design is shown in Figure \ref{fig:benchmark-design}. Four mixture conditions were created with differing amounts of human lysate, background lysate, and human spike-in ub-peptide mixture. Unmodified peptides from human lysate were used to estimate changes in global protein abundance. Background {\it E. coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate was treated as control and is expected to show no fold change in all comparisons except mix1-mix2 and mix3-mix4. The true log fold changes of spike-in modified peptides in each comparison can be seen in Figure \ref{fig:benchmark-design}(b). Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only.

@@

%\begin{table}[h!]
%\centering
%\begin{tabular}{| c | c |}
%\hline
% Comparison & Expected log fold change\\ [0.5ex]
% \hline\hline
% mix1-mix2 & -1\\
% \hline
% mix1-mix3 & 1\\
%\hline
% mix1-mix4 & 0\\
%\hline
% mix2-mix3 & -2\\
%\hline
% mix2-mix4 & -1\\
%\hline
% mix3-mix4 & 1\\
%\hline
%\end{tabular}
%\caption{The expected log fold change of modified spike-in peptides in Dataset 3.}
%\label{table:spikein_fold_change}
%\end{table}

@@

\medskip \noindent {\bf Experimental Design:} Luchetti et al. \cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. \todo{If this is a group comparison, explain why, even though the conditions are times. The table above says it is group comparison} The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with one LC-MS/MS run.

>>>>

Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of biomolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen,Bludau2022}. Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}.

@@

The technological difficulties in PTM identification and quantification challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. In particular, methods such as $t$-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~\cite{Ritchie_15a}. Such methods do not fully account for all the sources of uncertainty and variation in the experiment. As the result, they are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

The datasets in this manuscript both benchmarked the proposed approach in situations with known ground truth and represented a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real-world experimental settings with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, modifications, experimental designs and acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments.

@@

The simulation represented an idealistic case. 24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation. In each dataset, 1000 proteins had 10 unmodified features per protein. Each of the 1000 proteins had one PTM; each PTM was represented by 10 modified features. The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no difference between conditions. Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein. Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All differential PTMs were generated with an expected log fold change of 0.75 between conditions.

@@

@@

\medskip \noindent {\bf Experimental Design:} 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark. An overview of the experimental design is shown in Figure \ref{fig:benchmark-design}. Four mixture conditions were created with differing amounts of human lysate, background lysate, and human spike-in ub-peptide mixture. Unmodified peptides from human lysate were used to estimate changes in global protein abundance. Background {\it E. coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate is expected to change in the same way as the corresponding protein. Thus it is treated as control and is expected to show no fold change in all comparisons after adjusting for changes in the unmodified protein abundance. The true log fold changes of spike-in modified peptides in each comparison can be seen in Figure \ref{fig:benchmark-design}(b). Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only.

@@

%\begin{table}[h!]
%\centering
%\begin{tabular}{| c | c |}
%\hline
% Comparison & Expected log fold change\\ %[0.5ex]
% \hline\hline
% mix1-mix2 & -1\\
% \hline
% mix1-mix3 & 1\\
%\hline
% mix1-mix4 & 0\\
%\hline
% mix2-mix3 & -2\\
%\hline
% mix2-mix4 & 1\\
%\hline
% mix3-mix4 & -1\\
%\hline
%\end{tabular}
%\caption{The expected log fold change of %modified spike-in peptides in Dataset 3.}
%\label{table:spikein_fold_change}
%\end{table}

@@

\medskip \noindent {\bf Experimental Design:} Luchetti et al. \cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with one LC-MS/MS run.
